---
title: "Global Anemia Prevalence and Maternal Mortality Among Women"
subtitle: "Data Wrangling, Visualization and Reporting Wrap Up"
author: "Elvis Safary"
format:
  html:
    embeded-resources: true
    theme: lux
    toc: true
    toc-depth: 3
    code-fold: show
    code-copy: true
    smooth-scroll: true
    number-sections: false
    page-layout: article
    fontsize: 1.05em
    linestretch: 1.6
---

```{r include = FALSE}
# Set chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r}
#load package
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, DT, outbreaks, here, plotly, ggplot2, countrycode, scales, tinytex, bslib, htmltools)

#read data
anemia_raw <- read_csv(here("data/prevalence_anemia.csv"), skip_empty_rows = TRUE)

maternal_raw <- read_csv(here("data/maternal_mortality_ratio.csv"), skip_empty_rows = TRUE)

```


```{r}
# clean column names and reshape
anemia_long <- anemia_raw %>% 
  rename(country_code = geo, country = name) %>% 
  pivot_longer(cols = -c(country_code, country), #select all columns except these two
               names_to = "year",           # now column is 'year'
               values_to = "anemia_prev") %>% 
  mutate(year = as.integer(year),
         anemia_prev = as.numeric(anemia_prev))

#anemia_long
```

```{r}
maternal_long <- maternal_raw %>%
  rename(country_code = geo, country = name) %>%
  pivot_longer(cols = -c(country_code, country), 
               names_to = "year", 
               values_to = "mmr") %>%
  mutate(year = as.integer(year), 
         mmr = as.numeric(mmr))
#maternal_long
```

```{r}
# join two indicators
combined_long <- anemia_long %>%
  inner_join(maternal_long, by = c("country_code", "country", "year"))
#combined_long
```

```{r}
#keep year 2000 onwards
combined_long <- combined_long %>% filter(year >= 2000)
#combined_long
```

```{r}
# extract baseline (2000) and most recent (2023) for each country
baseline_data <- combined_long %>%
  filter(year == 2000) %>%
  select(country_code, country, anemia_2000 = anemia_prev, mmr_2000 = mmr)

latest_data <- combined_long %>%
  filter(year == 2023) %>%
  select(country_code, country, anemia_2023 = anemia_prev, mmr_2023 = mmr)
```

```{r}
# combine and compute changes
country_data <- baseline_data %>%
  inner_join(latest_data, by = c("country_code", "country")) %>%
  mutate(anemia_change = anemia_2023 - anemia_2000,
         mmr_change = mmr_2023 - mmr_2000,
         anemia_pct_change = (anemia_change / anemia_2000) * 100,
         mmr_pct_change = (mmr_change / mmr_2000) * 100,
         continent = countrycode(country_code, "iso3c", "continent")) %>% 
  mutate(continent = case_when(country_code == "SSD" ~ "Africa",
                               country_code == "KOS" ~ "Europe",
                               TRUE ~ continent))
```

```{r}
# Global averages for value boxes
global_anemia_2023 <- combined_long %>%
  filter(year == 2023) %>% summarise(avg = mean(anemia_prev, na.rm = TRUE)) %>% pull(avg)

global_mmr_2023 <- combined_long %>%
  filter(year == 2023) %>% summarise(avg = mean(mmr, na.rm = TRUE)) %>% pull(avg)
```


```{r}
layout_columns(
  col_widths = c(3, 3, 3, 3),   # four equal columns
  value_box(
    title = "Global Maternal Deaths in Conflict/Fragile Settings",
    value = "6 in 10",
    showcase = bsicons::bs_icon("exclamation-triangle"),
    theme = "danger",
    p("Countries affected by conflict or institutional fragility account for almost 60% of maternal deaths worldwide.")),
  value_box(
    title = "Maternal Mortality Ratio (2023)",
    value = paste0(round(global_mmr_2023, 0), " per 100,000"),
    showcase = bsicons::bs_icon("heart-pulse"),
    theme = "primary",
    p("Global average – down from 156 in 2000.")),
  value_box(
    title = "Anemia Prevalence (2023)",
    value = paste0(round(global_anemia_2023, 1), "%"),
    showcase = bsicons::bs_icon("droplet"),
    theme = "success",
    p("Women aged 15–49 years. Slight worsening since 2000.")),
  value_box(
    title = "Leading Cause of Maternal Death",
    value = "Haemorrhage (27%)",
    showcase = bsicons::bs_icon("droplet-half"),
    theme = "warning",
    p("Post‑partum bleeding remains the biggest direct killer.")))
```


# Gobal average anemia prevalence

-The global average anemia prevalence decreased from about 31% in 2000 to nearly 28% in 2023, indicating an improvement of women's nutritional and health status in many countries.

```{r}
# global average anemia per year
global_anemia <- combined_long %>%
  group_by(year) %>%
  summarise(global_avg = mean(anemia_prev, na.rm = TRUE))

anemia_plot <- ggplot(global_anemia, aes(x = year, y = global_avg)) +
  geom_line(color = "#2c3e50", size = 1.2) +
  geom_point(color = "#e74c3c", size = 2) +
  labs(title = "Global average anemia prevalence (2000–2023)",
       x = "Year", y = "Prevalence (%)") +
  theme_minimal()
ggplotly(anemia_plot)
```
# Distribution of Anemia prevalence in 2000 vs 2023
- The median anemia prevalence has decreased, and the interquartile range has widened, meaning inequality among countries might have increased.

```{r}
# create a combined dataframe for boxplot
anemia_compare <- country_data %>%
  select(country, anemia_2000, anemia_2023) %>%
  pivot_longer(cols = c(anemia_2000, anemia_2023),
               names_to = "year", values_to = "anemia") %>%
  mutate(year = if_else(year == "anemia_2000", 2000, 2023))

anemia_distr <- ggplot(anemia_compare, aes(x = factor(year), y = anemia, fill = factor(year))) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Distribution of anemia prevalence in 2000 vs 2023",
       x = "Year", y = "Anemia prevalence (%)") +
  theme_minimal() +
  scale_fill_manual(values = c("#3498db", "#e74c3c"), guide = "none")
ggplotly(anemia_distr)
```
# Average Maternal Mortality ratio from 2000 to 2023
- Maternal mortality has declined globally, from about 156 per 100,000 in 2000 to roughly 96 in 2023, a substantial improvement.

```{r}
global_mmr <- combined_long %>%
  group_by(year) %>%
  summarise(global_avg = mean(mmr, na.rm = TRUE))

mmr_plot <- ggplot(global_mmr, aes(x = year, y = global_avg)) +
  geom_line(color = "#27ae60", size = 1.2) +
  geom_point(color = "#c0392b", size = 2) +
  labs(title = "Global average maternal mortality ratio (2000–2023)",
       x = "Year", y = "Deaths per 100,000 live births") +
  theme_minimal()
ggplotly(mmr_plot)
```
# mmr by continent

```{r}
region_mmr <- combined_long %>%
  filter(year == 2023) %>%
  left_join(select(country_data, country_code, continent), by = "country_code") %>%
  group_by(continent) %>%
  summarise(avg_mmr = mean(mmr, na.rm = TRUE)) %>%
  filter(!is.na(continent))

p_region <- ggplot(region_mmr, aes(x = reorder(continent, -avg_mmr), y = avg_mmr, fill = continent)) +
  geom_col() +
  labs(title = "Maternal mortality ratio by continent (2023)",
       x = NULL, y = "Deaths per 100,000 live births") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")
ggplotly(p_region)
```

# Relationship between Anemia and MMR
- There is a weak positive correlation. Countries that reduced MMR also tended to reduce anemia prevalence, but many outliers exist. For example, some countries saw anemia rise despite stable MMR.

```{r}
# relationship between change in anemia and mmr

rship_plot <- ggplot(country_data, aes(x = mmr_change, y = anemia_change)) +
  geom_point(alpha = 0.6, color = "#2c3e50", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#e74c3c", linetype = "dashed") +
  labs(title = "Relationship between changes in MMR and anemia (2000–2023)",
       x = "Change in maternal mortality ratio (deaths per 100,000)",
       y = "Change in anemia prevalence (percentage points)") +
  theme_minimal()
ggplotly(rship_plot)
```

# causes of maternal death

```{r}
# Data from the WHO infographic
causes <- tibble(
  cause = c("Haemorrhage", "Communicable & non‑communicable diseases",
            "Hypertensive disorders", "Abortion / non‑birth outcomes", "Sepsis"),
  percentage = c(27, 23, 16, 8.4, 6.6))

maternal_causes <- plot_ly(causes, labels = ~cause, values = ~percentage, type = 'pie',
                    textposition = 'inside', textinfo = 'percent+label',
                    marker = list(colors = c("#c0392b", "#e67e22", "#f1c40f", "#2ecc71", "#3498db"))) %>%
  layout(title = "Leading causes of maternal death (global estimate)",
         showlegend = FALSE)
maternal_causes
```


# Comparison Table between Anemia and MMR

- This interactive table displays a side‑by‑side comparison of anemia prevalence and maternal mortality ratio (MMR) for each country, covering the years 2000 and 2023. Additional columns include absolute changes for both indicators, allowing users to sort, search, and identify countries that have improved or worsened. The colour‑coded bars visually highlight the magnitude of change.


```{r}
summary_table <- country_data %>%
  select(country, anemia_2000, anemia_2023, anemia_change,
         mmr_2000, mmr_2023, mmr_change) %>%
  arrange(desc(anemia_change)) %>%
  mutate(across(where(is.numeric), ~ round(., 1)))

datatable(summary_table,
          caption = "Full country comparison: anemia and MMR changes (2000 to 2023)",
          options = list(pageLength = 15, scrollX = TRUE),
          rownames = FALSE) %>%
  formatStyle("anemia_change",
              background = styleColorBar(range(summary_table$anemia_change, na.rm = TRUE), "lightpink"),
              backgroundSize = "100% 90%",
              backgroundRepeat = "no-repeat") %>%
  formatStyle("mmr_change",
              background = styleColorBar(range(summary_table$mmr_change, na.rm = TRUE), "lightgreen"),
              backgroundSize = "100% 90%",
              backgroundRepeat = "no-repeat")
```

# Key takeaways

- Maternal mortality has fallen globally, however anemia has slightly worsened, and regional gaps remain wide.
- Anemia and MMR are weakly correlated – Countries that improve one often improve the other, but exceptions highlight the need for targeted nutrition and health‑system interventions.
- Data empowers action – Transparent, interactive dashboards help policymakers identify high‑burden countries and allocate resources effectively.




